preciseTAD: A machine learning framework for precise TAD boundary prediction at base pair resolution

### 2020-07-31 Manuscript/manuscript.Rmd
editted results

Using our optimally built random forest model trained on ARROWHEAD/Peakachu boundaries, we attempted to predict TAD/loop boundaries at base resolution. Intuitively, the predictor-response space is built on a base-resolution level, and the model trained on a bin-level space predicts the probability of each base being a boundary. Our method, referred to as *preciseTAD*, leverages a user-specified [??? We need pre-trained models and understanding how they perform on different cell lines;!!!working on it] model with density-based hierarchical clustering (DBSCAN) and a scalable partitioning around mediods clustering (CLARA) to predict the boundary location ad base resolution (see Methods).

When built using ARROWHEAD ground truth boundaries at 5 kb, *preciseTAD* was able to predict a total of ??? and ??? boundaries for GM12878 and K562 cell lines respectively. We reported less predicted TAD boundaries than ARROWHEAD (Table 1, Supplementary Table 2) [??? Where is this info?;!!!talk about in meeting]. This can be attributed to ARROWHEAD's boundaries frequently being separated by gaps, the effect especially pronounced near sparse areas of Hi-C contact matrices (Figure 5A) [??? I don't see it, data not shown?]. However, under the framework of *preciseTAD*, any two TADs are inherently conjoined by only one predicted TAD boundary, leading to comparatively less TAD boundaries. [??? Unclear explanation, so what? preciseTAD is driven by genomic annotations. May it be too stringent thresholds? This all may be for the discussion, if we have a good understanding why] We also found that predicted TAD boundaries identified by *preciseTAD* were visually more likely to demarcate high concentrations of inter-chromosomal contacts as seen on a Hi-C contact matrix (Figure 5A). [???add peakachu totals when ready]

Next, we evaluated the biological significance of *preciseTAD* boundaries. Signal heatmaps [??? Make sure to mention DeepTools in Methods;!!!added] indicated that our predicted boundaries were more enriched for CTCF sites than either ARROWHEAD or Peakachu boundaries alone (Figure 5 B-C). Likewise, *preciseTAD* boundaries were also more enriched for other molecular drivers of 3D chromatin including RAD21, SMC3, and ZNF143 for both GM12878 and K562 cell lines (Supplementary Figures 8-9). When compared to called TAD or loop boundaries, *preciseTAD* boundaries were found to be statistically significantly closer to the same cell line-specific top ranked TFBS (p=??? and p=??? vs ARROWHEAD and Peakachu respectively) (Figure 5D, Supplementary Figure 10-11). We also inspected the effect of resolution on locations of our predicted TAD boundaries. We found that *preciseTAD* boundaries exhibited higher pairwise Jaccard indices between adjacent resolutions than ARROWHEAD boundaries, thus implying stronger invariance toward resolution (Figure 5E).

@@

## *preciseTAD* boundaries are conserved across cell lines

TAD boundaries are known to be relatively conserved across cell lines [@dixon2012topological; @krefting2018evolutionary; @nora2012spatial; @dixon2015chromatin; @rao20143d; @Schmitt2016aa]. To assess conservation, we evaluated the overlap between cell-line-specific boundaries detected by *preciseTAD*, ARROWHEAD, and Peakachu after flanking them by the resolution of the Hi-C data. Boundaries involving ARROWHEAD were flanked by 5 kb, while boundaries involving Peakachu were flanked by 10 kb. We found that a significant proportion of flanked boundaries predicted by *preciseTAD* overlapped between GM12878 and K562. Specifically, 45%??? and 61%??? of *preciseTAD* boundaries were conserved across cell lines when built on ARROWHEAD and Peakachu boundaries respectively (Figure 6A). Additionally, the median Jaccard index (MJI) was found to be 0.472 (IQR=0.441-0.503; p=???) and 0.647 (IQR=0.646-0.647; p=???) for *preciseTAD* predicted TAD and loop boundaries (Supplementary Table 5). Comparatively, only 25% (MJI=0.284; IQR=0.268-0.301) and 49% (MJI=0.509; IQR=0.503-0.515) of called/predicted ARROWHEAD and Peakachu boundaries were conserved (Figure 6A, Supplementary Table 5). These results demonstrate that boundaries predicted using genomic annotations (*preciseTAD*) are better conserved across cell lines.

The majority of boundaries predicted by the ARROWHEAD-trained *preciseTAD* model represented a subset of the larger group of boundaries predicted by the Peakachu-trained model (Figure 6B). This is expected as loop boundaries (Peackachu) are more abundant, while TAD boundaries (ARROWHEAD) represent the narrower level of the 3D chromatin organization. 83% and 75% of predicted TAD and loop boundaries overlapped with each other for GM12878 and K562 cell lines, respectively (Figure 6B). Together with the results shown on Figure 4, the high overlap between boundaries predicted by ARROWHEAD- and Peakachu-trained models suggests that TAD and loop boundaries may be driven by similar protein complexes.

>>>>

Using our optimally built random forest model trained on ARROWHEAD/Peakachu boundaries, we attempted to predict TAD/loop boundaries at base resolution. Intuitively, the predictor-response space is built on a base-resolution level, and the model trained (using the optimal characteristics outlined above) on a bin-level space predicts the probability of each base being a boundary. Our method, referred to as *preciseTAD*, leverages a user-specified model with density-based hierarchical clustering (DBSCAN) and a scalable partitioning around mediods clustering (CLARA) to predict the boundary location at base resolution (see Methods).

When built using ARROWHEAD and PEAKACHU ground truth boundaries at 5 kb and 10 kb respectively, *preciseTAD* was able to predict a total of 7846 TAD and 15838 chromatin loop boundaries on GM12878, as well as 8883 TAD and 11016 chromatin loop boundaries on K562. Noticeably, we reported less predicted TAD boundaries at 5 kb than ARROWHEAD on both cell lines (Table 1, Supplementary Table 2). This can be attributed to ARROWHEAD's inflation of called TADs, especially at higher resolutions, that do not correspond to intra-chromosomal contacts featured on a contact matrix (Figure 1A). Likewise, ARROWHEAD often produces nested and overlapped TADs that do not have demarcated boundaries, further exacerbating this inflation. However, under the framework of *preciseTAD*, any two TADs are inherently conjoined by only one predicted TAD boundary, leading to comparatively less TAD boundaries. [??? Unclear explanation, so what? preciseTAD is driven by genomic annotations. May it be too stringent thresholds? This all may be for the discussion, if we have a good understanding why] Additionally, *preciseTAD* boundaries were visually more likely to demarcate high concentrations of intra-chromosomal contacts as seen on a Hi-C contact matrix (Figure 5A). *preciseTAD* also predicted fewer chromatin loop boundaries than PEAKACHU (Table 1, Supplementary Table 2). This can be contributed to the fact that *preciseTAD* leverages 4 known drivers of 3D chromatin including CTCF, RAD21, SMC3, and ZNF143, contrasting PEAKACHU which considers only CTCF sites. This methodology likely leads to many more false positives, resulting in inflated number of chromatin loops.

Next, we evaluated the biological significance of *preciseTAD* boundaries. Signal heatmaps indicated that our predicted boundaries were more enriched for CTCF sites than either ARROWHEAD or PEAKACHU boundaries alone (Figure 5 B-C). Likewise, *preciseTAD* boundaries were also more enriched for other molecular drivers of 3D chromatin including RAD21, SMC3, and ZNF143 for both GM12878 and K562 cell lines (Supplementary Figures 5-6). When compared to called TAD or loop boundaries, *preciseTAD* boundaries were found to be statistically significantly closer to the same cell line-specific top ranked TFBS (p<0.001 and p<0.001 versus ARROWHEAD and PEAKACHU respectively) (Figure 5D, Supplementary Figure 7).

## *preciseTAD* boundaries are more conserved across cell lines

TAD boundaries are known to be relatively conserved across cell lines [@dixon2012topological; @krefting2018evolutionary; @nora2012spatial; @dixon2015chromatin; @rao20143d; @Schmitt2016aa]. To assess conservation, we evaluated the overlap between cell line-specific boundaries detected by *preciseTAD*, ARROWHEAD, and Peakachu after flanking them by the resolution of the Hi-C data. Boundaries involving ARROWHEAD were flanked by 5 kb, while boundaries involving PEAKACHU were flanked by 10 kb. We found that 35% of *preciseTAD* predicted TAD boundaries and 55% of *preciseTAD* predicted loop boundaries were were conserved between cell lines, compared to only 26% and 55% for ARROWHEAD and PEAKACHU alone (Figure 5E) [???should I mention jaccard or scrap it? significantly different for arrowhead but not peakachu].

<!--Additionally, the median Jaccard index (MJI) was found to be 0.472 (IQR=0.441-0.503; p=???) and 0.647 (IQR=0.646-0.647; p=???) for *preciseTAD* predicted TAD and loop boundaries (Supplementary Table 5). Comparatively, only 25% (MJI=0.284; IQR=0.268-0.301) and 49% (MJI=0.509; IQR=0.503-0.515) of called/predicted ARROWHEAD and Peakachu boundaries were conserved (Figure 6A, Supplementary Table 5). These results demonstrate that boundaries predicted using genomic annotations (*preciseTAD*) are better conserved across cell lines.-->

<!--We also inspected the effect of resolution on locations of our predicted TAD boundaries. We found that *preciseTAD* boundaries exhibited higher pairwise Jaccard indices between adjacent resolutions than ARROWHEAD boundaries, thus implying stronger invariance toward resolution (Figure 5E).-->

@@

## *preciseTAD* predicted TAD boundaries are nested within predicted loop boundaries

The majority of boundaries predicted by the ARROWHEAD-trained *preciseTAD* model represented a subset of the larger group of boundaries predicted by the Peakachu-trained model. We found that 95.6% and 95.7% of our predicted TAD boundaries were overlapped by predicted loop boundaries for GM12878 and K562, respectively (Supplementary Figure 8). This is expected as loop boundaries (Peackachu) are more abundant, while TAD boundaries (ARROWHEAD) represent the narrower level of the 3D chromatin organization. Together with the results shown on Figure 4, the high overlap between boundaries predicted by ARROWHEAD- and Peakachu-trained models suggests that TAD and loop boundaries may be driven by similar protein complexes.
